Interpolation of Sparse High-Dimensional Data\thanks {This work was supported by the National Science Foundation Grant CNS-1565314.}

### 2020-04-01 [2019-02]_NUMA_(interpolant_error_bound)/paper.tex
Finished first draft of response to reviewers for the NA journal paper.

% Insert here the call for the packages your document requires
\usepackage{graphicx}
\usepackage{tikz}     % For adding axes to plots
\usetikzlibrary{positioning}
\usepackage{graphicx} % For including PDFs
\usepackage{amsmath}  % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{booktabs} % For formal tables
\usepackage{tabularx} % For tables with nice column widths.
\usepackage[colorlinks,allcolors=blue]{hyperref}
\usepackage{float}    % For controlling the placement of tables.

%% -------------------------------------------------------------------
%% Command for changing the color of text within a block.
\usepackage{xcolor}   % Color package for coloring text.
\usepackage[normalem]{ulem} % For strikethrough text and the \sout{} command.
\usepackage[xcolor,rightbars]{changebar}% Package for adding bars into margin next to comments.
\renewcommand{\changebarsep}{5mm}   % Push the bar farther into the margin.
\renewcommand{\changebarwidth}{1mm} % Increase the width of the bar.
\cbcolor{red}
\newcommand{\changed}[2]{\cbstart{\color{red}#1}{\color{gray}\sout{#2}}\cbend{}}
%% -------------------------------------------------------------------

@@

\title{Interpolation of Sparse High-Dimensional Data\thanks
  {This work was supported by the National Science Foundation Grant CNS-1565314.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}
%\titlerunning{Short form of title}        % if too long for running head

@@

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

@@

Regression and interpolation have a considerable theoretical base in
one dimension \cite{cheney2009course}. \changed{Common techniques
  include Lagrangian interpolation, piecewise linear interpolation,
  and more generally spline interpolation via B-splines
  \cite{de1978practical}.}{Splines in particular are well understood
  as an interpolation technique in one dimension
  \cite{de1978practical}, particularly B-splines}. Tensor products of
B-splines \cite{unther1996interpolating} or other basis functions have
an unfortunate exponential scaling with increasing
dimension. Exponential scaling prohibits tensor products from being
reasonably applied beyond three-dimensional data. In order to address
this dimensional scaling challenge, C. de Boor and others proposed box
splines \cite{de2013box}, of which one of the approximation techniques
in this work is composed \cite{lux2018novel}.

The theoretical foundation of low dimensional interpolation allows the
construction of strong error bounds that are absent from high
dimensional problems. This work extends some known results regarding
the secant method \cite{dennis1996numerical} to construct an
interpolation error bound for problems of arbitrary dimension. These
error bounds are useful, considering the same cannot be said for
regression algorithms in general. The maximum complexity of an
interpolant is bounded by the amount of data available, while the
maximum complexity of a regressor is bounded by both the amount of
data and the chosen parametric form. For this reason, generic uniform
bounds are largely unobtainable for regression techniques on arbitrary
approximation problems, even when the approximation domain is bounded.

@@

The most challenging problem when scaling in dimension is that the
number of possible interactions between dimensions grows
exponentially. Quantifying all possible interactions becomes
intractable, and hence beyond three-dimensional data mostly linear
models are used. That is not to say nonlinear models are absent, but
nonlinearities are often either preconceived or model pairwise
interactions between dimensions at most. Even globally nonlinear
approximations such as neural networks are constructed from
compositions of summed low-interaction functions
\cite{clevert2015fast}.

@@

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k - \bigl\|x -
x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x - x^{(k)}\bigr\|_2 \bigr)\bigr)^2$}{}
and $r_k \in \mathbb{R}$ is the smallest radius such that at least
$d+1$ other points $x^{(j)}$, $j \not = k$, are inside the closed
Euclidean ball of radius $r_k$ about $x^{(k)}$. The computational
complexity of ShepMod is $\mathcal{O}(n^2d)$. This paper uses a
Fortran 95 implementation of ShepMod derived from SHEPPACK
\cite{thacker2010algorithm}.

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
  \caption{\color{red} On the left above is a depiction of the radius
    of influence for three chosen points of a collection using the
    modified Shepard criteria. On the right is the shape of the weight
    function $W_k$ for all data points, where $W_k(x) = 0$ when
    $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen that in the limit
    as $W_k$ goes to infinity at $x_k$, the relative weight of $x_k$
    goes to one.
  \vspace{-.1cm}}
  \label{fig:shepard}
\end{figure}
}{}

@@

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth,]{VM_Construction.pdf}
  \caption{\color{red} Above is a depiction of the Voronoi cell
    boundaries (dashed lines) about a set of interpolation points
    (dots). In this example, the Voronoi model basis function about
    the center most point has nonzero weight in the shaded region and
    transitions from a value of one at the point to zero at the
    boundary of the twice expanded Voronoi cell (solid line).
  \vspace{-.1cm}}
  \label{fig:voronoi-basis}
\end{figure}
}{}

@@

\begin{figure}[htb]
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen. The Delaunay
    method chose three source distributions (dotted lines) and
    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
    weighted sum of the three known CDFs produces the predicted CDF
    (dashed line). The KS Statistic (arrow) computed between the true
    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
    example. For this example the KS test null hypothesis is rejected
    at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

@@

The two-sample KS test is a useful nonparametric test for comparing
two CDFs while only assuming stationarity, finite mean, and finite
variance. The null hypothesis (that two CDFs come from the same
underlying distribution) is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
applications of the KS test presented in this work $n_1 = n_2$. An
example of the \changed{end-to-end}{round-trip} prediction methodology from known and
predicted distributions to the calculation of error can be seen in
Figure \ref{fig:prediction-example}. A brief listing of relevant
statistical terms used throughout this work is provided in
the Appendix (Section \ref{sec:appendix}).

@@

\begin{proofdot}
  \def\g{\gamma_g}
  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
  = \gamma \|x_j-x_0\|_2^2$.

  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
  and so $g'(0) - \g z \le g'(z) = 0$. Now
  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
  \begin{align*}
    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
  \end{align*}
  \noindent a contradiction for the value of $g(\tilde t)$. A similar
  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
  \g/2$. In terms of $f$,
  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
  \implies \|\nabla f(x_0)-c\|_2 \le \frac{\sqrt{d} \, \gamma
    k^2}{2\sigma_d}.$$ \qed
\end{proofdot}

@@

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate.  Only linear convergence to the true function can
be achieved in practice, without the incorporation of additional
observations. Notice that the approximation error is largely
determined by the spacing of observed data. Predictions made by
simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher
error.

@@

The error of a linear interpolant constructed from well spaced points
is dominated by the distance to the nearest data point. In order to
uniformly decrease the distance to the nearest data point across the
unit cube, exponentially more data points must be available for
approximation. For this experiment $N = 2(4^i),$ $i = 1,$ $\ldots,$
$6,$ chosen points are kept well spaced by computing approximate
Fekete points. Fekete points have a history in potential theory
\cite{kovari_pommerenke_1968} and are most generally defined as those
points that maximize the absolute value of a Vandermonde
determinant. Here, the QR method outlined in \cite{bos2010computing}
is used to identify approximate Fekete points from a Vandermonde
matrix. A multivariate polynomial basis of size $2N$ is constructed
containing all polynomials of degree $\leq n$ for the largest $n$ such
that ${d+n \choose n} \leq 2N$ and $2N - {d+n \choose n}$
arbitrarily selected polynomials of degree $n + 1.$ The Vandermonde
matrix for these $2N$ polynomials is used to select $N$ approximate
Fekete points.

Finally an additional aspect is added to this test problem by
incorporating random uniform noise into the evaluations of $f.$ For
each test two experiments are executed, one with exact function
evaluations (an interpolation problem) and one with a constant
signal-to-noise (SNR) ratio of 10:1 (a regression problem).

@@

The bound from the theorem suggests that exponentially increasing the
number of \changed{well-spaced}{} data points should result in a
decreasing error. The $d = 2$ test seen on the left half of Figure
\ref{fig:convergence-2d} shows a consistent decrease in error for
Delaunay and also shows the eventual accuracy plateau obtained by a
parametric regression form (the MLP, at roughly 500 points). On the
right hand side of Figure \ref{fig:convergence-2d} the random noise
clearly prohibits Delaunay from converging to $f$, while the MLP is
able to improve its approximation with more data points on average.
The convergence result for a very low-dimensional problem like this is
expected.  However, intuition fails for higher dimensional problems.

@@

\changed{
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.9\textwidth,]{theorem-convergence.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Euclidean Distance};
  \end{tikzpicture}
  \caption{\color{red}The distribution of absolute error, distance to
    the nearest data point, smallest singular value and the longest
    edge of the simplex containing each approximation point in the
    tests from Fig. \ref{fig:convergence-2d} and
    Fig. \ref{fig:convergence-20d} for Delaunay. In two dimensions it
    can be seen that $\|z - x_0\|_2$, $\sigma_d$, and $k$ all shrink
    at the same rate for well-spaced approximation points, resulting
    in a faster rate of decrease for approximation error. Notice that
    in higher dimension the data remains sparse even with thousands of
    data points. This is seen by the relatively small reduction in
    $k$, and this explains the slow reduction in error seen by
    Delaunay in Fig. \ref{fig:convergence-20d}.}
  \label{fig:data-spacing}
\end{figure}
}{}

\changed{The results achieved for both the $d = 2$ and $d = 20$ align
  with the theoretical error bound as can be seen in Figure
  \ref{fig:data-spacing}. In low dimension, thousands of data points
  meaningfully reduce the sparsity of the approximation problem.
  However, in higher dimension it takes many more points to achieve a
  reduction in data sparsity. The same result holds for a Latin
  hypercube well-spaced point set, suggesting that sparsity is an
  inherent problem in high dimension approximation problems. This
  result is not a product of the chosen test function, but rather the
  exponential increase in complexity that accompanies increased
  dimension.}{}

In summary, the regime of interpolation is \changed{}{much} greater
for moderate to high dimensional problems \changed{due to increased
  sparsity}{}, even for functions with random noise incorporated into
their evaluation. Acknowledging the viability of interpolation for
problems of moderate dimension, the next section will consider
real-world problems of similar proportion (thousands of examples in
tens of dimensions).

@@

In addition to the figures displaying approximation results for each
data set, tables of accompanying numerical results are located in the
Appendix (Section \ref{sec:appendix}). All of the test data sets
capture underlying functions that are almost certainly stochastic. As
described in Section \ref{sec:introduction}, regression techniques
appear most appropriate for these problems. However, typically data
grows exponentially more sparse with increasing dimension. Given that
sparse data regressors tend towards interpolation and, as demonstrated
in Section \ref{sec:analytic}, interpolants produce similar (if not
identical) results, it is presumed that interpolants are equally
viable approximation techniques on these problems.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of forest fire area burned under recorded weather
    conditions. The data is presented on a $\ln$ scale because
    most values are small with exponentially fewer fires on record
    that burn large areas.}
  \label{fig:hist-forest-fire}
\end{figure}

>>>>

% Insert here the call for the packages your document requires
\usepackage{graphicx}
\usepackage{tikz}     % For adding axes to plots
\usetikzlibrary{positioning}
\usepackage{graphicx} % For including PDFs
\usepackage{amsmath}  % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{booktabs} % For formal tables
\usepackage{tabularx} % For tables with nice column widths.
\usepackage[colorlinks,allcolors=blue]{hyperref}
\usepackage{float}    % For controlling the placement of tables.
\usepackage{enumitem} % For changing item indents in enumerate.

%% -------------------------------------------------------------------
%% Command for changing the color of text within a block.
\usepackage{xcolor}   % Color package for coloring text.
\usepackage[normalem]{ulem} % For strikethrough text and the \sout{} command.
\usepackage[xcolor,rightbars]{changebar}% Package for adding bars into margin next to comments.
\renewcommand{\changebarsep}{5mm}   % Push the bar farther into the margin.
\renewcommand{\changebarwidth}{1mm} % Increase the width of the bar.
\cbcolor{red}
%% \newcommand{\changed}[2]{\cbstart{\color{red}#1}{\color{gray}\sout{#2}}\cbend{}}
\newcommand{\changed}[2]{\cbstart{\color{red}#1}{}\cbend{}}
%% -------------------------------------------------------------------

@@

%% Include the response to reviewers at the beginning of the document.
\include{tchlux-NA-response-2020-04}

%% Set the title.
\title{Interpolation of Sparse High-Dimensional Data\thanks
  {This work was supported by the National Science Foundation Grant CNS-1565314.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}
%\titlerunning{Short form of title}        % if too long for running head

@@

\date{Received: date / Accepted: date}
%% The correct dates will be entered by the editor

{\let\newpage\relax\maketitle}

%% \maketitle

\setcounter{page}{1}

@@

Regression and interpolation have a considerable theoretical base in
one dimension \cite{cheney2009course}. \changed{Common techniques
  include Lagrangian interpolation, piecewise linear interpolation,
  and more generally spline interpolation via B-splines
  \cite{de1978practical}}{Splines in particular are well understood
  as an interpolation technique in one dimension
  \cite{de1978practical}, particularly B-splines}. Tensor products of
B-splines \cite{unther1996interpolating} or other basis functions have
an unfortunate exponential scaling with increasing
dimension. Exponential scaling prohibits tensor products from being
reasonably applied beyond three-dimensional data. In order to address
this dimensional scaling challenge, C. de Boor and others proposed box
splines \cite{de2013box}, of which one of the approximation techniques
in this work is composed \cite{lux2018novel}.

The theoretical foundation of low dimensional interpolation allows the
construction of strong error bounds that are \changed{lacking
  in}{absent from} high dimensional problems. \changed{A large body of
  literature exists studying the construction and application of
  sparse grids \cite{bungartz2004sparse,chkifa2014high}, which attempt
  to address the problem of dimensional scaling. The sparse grid point
  sets combined with specific approximation functions are used to
  construct total error bounds by considering projections onto
  orthogonal polynomials
  \cite{migliorati2013approximation,nobile2016convergence}. However,
  sparse grids have exponential size in dimension making them
  intractable for more than tens of dimensions and do not immediately
  apply to the problem of scattered data interpolation.}{} This work
extends some known results regarding the secant method
\cite{dennis1996numerical} to construct \changed{a uniform error bound
  for interpolants over arbitrary scattered data in any dimension}{an
  interpolation error bound for problems of arbitrary
  dimension}. These error bounds are useful, considering
\changed{similar uniform bounds}{the same cannot} cannot be
\changed{constructed}{said} for regression algorithms in general. The
maximum complexity of an interpolant is bounded by the amount of data
available, while the maximum complexity of a regressor is bounded by
both the amount of data and the chosen parametric form. For this
reason, generic uniform bounds are largely unobtainable for regression
techniques on arbitrary approximation problems, even when the
approximation domain is bounded.

@@

The most challenging problem when scaling in dimension is that the
number of possible interactions between dimensions grows
exponentially. Quantifying all possible interactions becomes
intractable, and hence \changed{problems}{} beyond \changed{ten to
  twenty dimensions usually rely on linear models}{three-dimensional
  data mostly linear models are used}. That is not to say nonlinear
models are absent, but nonlinearities are often either preconceived or
model \changed{interactions between small subsets of}{pairwise
  interactions between} dimensions at most. Even globally nonlinear
approximations such as neural networks are constructed from
compositions of summed low-interaction functions
\cite{clevert2015fast}.

@@

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k -
  \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x -
  x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x \neq x^{(k)}$,}{} and $r_k
\in \mathbb{R}$ is the smallest radius such that at least $d+1$ other
points $x^{(j)}$, $j \not = k$, are inside the closed Euclidean ball
of radius $r_k$ about $x^{(k)}$. The computational complexity of
ShepMod is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95
implementation of ShepMod derived from SHEPPACK
\cite{thacker2010algorithm}.

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
  \caption{\color{red} On the left above is a depiction of the radius
    of influence for three chosen points of a collection in two
    dimensions using the modified Shepard criteria. On the right is
    the shape of the weight function $W_k$ for all data points, where
    $W_k(x) = 0$ when $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen
    that in the limit as $W_k$ tends towards infinity at $x_k$, the
    relative weight of $x_k$ tends to one.
  \vspace{-.1cm}}
  \label{fig:shepard}
\end{figure}
}{}

@@

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth,]{VM_Construction.pdf}
  \caption{\color{red} Above is a depiction of the Voronoi cell
    boundaries (dashed lines) about a set of interpolation points
    (dots) in two dimensions. In this example, the Voronoi model basis
    function about the center most point has nonzero weight in the
    shaded region and transitions from a value of one at the point to
    zero at the boundary of the twice expanded Voronoi cell (solid
    line).
  \vspace{-.1cm}}
  \label{fig:voronoi-basis}
\end{figure}
}{}

@@

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth,]{example_distribution_prediction.pdf}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen\color{red},
    where M means $\times 10^6$\color{black}. The Delaunay method
    chose three source distributions (dotted lines) and assigned
    weights \color{red}\{.1, .3, .6\} \color{black} (top to bottom at
    \color{red}middle\color{black}). The weighted sum of the three
    known CDFs produces the predicted CDF (dashed line). The KS
    Statistic (\color{red}vertical line\color{black}) computed between
    the true CDF (solid line) and predicted CDF (dashed line) is 0.2
    for this example. For this example the KS test null hypothesis is
    rejected at $p$-value 0.01, however it is not rejected at
    $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}
}{}

@@

The two-sample KS test is a useful nonparametric test for comparing
two \changed{empirical}{} CDFs while only assuming stationarity,
finite mean, and finite variance. The null hypothesis (that two
\changed{empirical}{} CDFs come from the same underlying distribution)
is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)}
\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$ with distribution sample
sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test
presented in this work $n_1 = n_2$. An example of \changed{the process
  of generating predicted distributions from known distributions and
  the subsequent}{the round-trip prediction methodology from known and
  predicted distributions to the} calculation of error can be seen in
Figure \ref{fig:prediction-example}. A brief listing of relevant
statistical terms used throughout this work is provided in the
Appendix (Section \ref{sec:appendix}).

@@

\begin{proofdot}
  \def\g{\gamma_g}
  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
  = \gamma \|x_j-x_0\|_2^2$. \changed{The following is visualized in 
  Figure \ref{fig:lemma3}.}{}

  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
  and so $g'(0) - \g z \le g'(z) = 0$. Now \changed{by integrating}{}
  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
  \changed{acknowledging $g'(0) > \g / 2$}{}
  \begin{align*}
    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
  \end{align*}
  \noindent a contradiction for the value of $g(\tilde t)$. A similar
  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
  \g/2$. In terms of $f$,
  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
  \implies \|\nabla f(x_0)-c\|_2 \le \frac{\sqrt{d} \, \gamma
    k^2}{2\sigma_d}.$$ \qed
\end{proofdot}

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{example_lemma_3.pdf}
  \def\g{\gamma_g}
  \caption{\color{red} Three different scenarios visualizing {\it
      Lemma \ref{lemma:3}}, where $g(t)$ is the difference between a
    piecewise linear interpolant and the approximated function along a
    normalized line segment between interpolation points, $g'(t)$ is
    $\g$-Lipschitz continuous, and $w$ and $\tilde t$ are defined in
    the proof. Leftmost is a randomly chosen permissible shape of $g$
    and $g'$. The middle is the only possible shape of $g$ and $g'$
    given $g'(0) = \g / 2$, establishing the case of equality in the
    lemma. Rightmost is the resulting contradiction when $g'(0) > \g /
    2$, notice it is impossible to ensure $g'(t)$ is $\g$-Lipschitz
    continuous and satisfy $g(1) = 0$ (highlighted with red circle on
    the right).
  \vspace{-.1cm}}
  \label{fig:lemma3}
\end{figure}
}{}

@@

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate. Only linear convergence to the true function
\changed{is}{can be} achieved in practice\changed{}{,} without the
incorporation of additional observations\changed{, by moving closer to
  interpolation points}{}. Notice that the approximation error is
largely determined by the spacing of observed data. Predictions made
by simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher error.

@@

The error of a linear interpolant constructed from well spaced points
is dominated by the distance to the nearest data point. In order to
uniformly decrease the distance to the nearest data point across the
unit cube, exponentially more data points must be available for
approximation. For this experiment $N = 2(4^i),$ $i = 1,$ $\ldots,$
$6,$ chosen points are kept well spaced by computing approximate
Fekete points. \changed{The following experiment was also run with a
  Latin hypercube design \cite{park1994optimal}, which produced almost
  identical results that are not reported here.}{} Fekete points have
a history in potential theory \cite{kovari_pommerenke_1968} and are
most generally defined as those points that maximize the absolute
value of a Vandermonde determinant. Here, the QR method outlined in
\cite{bos2010computing} is used to identify approximate Fekete points
from a Vandermonde matrix. A multivariate polynomial basis of size
$2N$ is constructed containing all polynomials of degree $\leq n$ for
the largest $n$ such that ${d+n \choose n} \leq 2N$ and $2N - {d+n
  \choose n}$ arbitrarily selected polynomials of degree $n + 1.$ The
Vandermonde matrix for these $2N$ polynomials is used to select $N$
approximate Fekete points.

Finally an additional aspect is added to this test problem by
incorporating random uniform noise into the evaluations of $f.$ For
each test two experiments are executed, one with exact function
evaluations (\changed{an interpolation problem}{}) and one with a
constant signal-to-noise ratio (SNR) of 10:1 (\changed{a regression
  problem}{}).

@@

The bound from the theorem suggests that \changed{by}{exponentially}
increasing the number of \changed{well-spaced}{} data points,
\changed{approximation error can be reliably decreased}{results in a
  decreasing error}. The $d = 2$ test seen on the left half of Figure
\ref{fig:convergence-2d} shows a consistent decrease in error for
Delaunay and also shows the eventual accuracy plateau obtained by a
parametric regression form (the MLP, at roughly 500 points). On the
right hand side of Figure \ref{fig:convergence-2d} the random noise
clearly prohibits Delaunay from converging to $f$, while the MLP is
able to improve its approximation with more data points on average.
The convergence result for a very low-dimensional problem like this is
expected.  However, intuition fails for higher dimensional problems.

@@

\changed{
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.9\textwidth,]{theorem-convergence.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Error, Distance, and Smallest SV };
  \end{tikzpicture}
  \caption{\color{red}The distribution of absolute error, distance to
    the nearest data point, smallest singular value (SV) and the
    longest edge of the simplex containing each approximation point in
    the tests from Fig. \ref{fig:convergence-2d} and
    Fig. \ref{fig:convergence-20d} for Delaunay. In two dimensions it
    can be seen that $\|z - x_0\|_2$, $\sigma_d$, and $k$ all shrink
    at the same rate for well-spaced approximation points, resulting
    in a faster rate of decrease for approximation error. Notice that
    in higher dimension the data remains sparse even with thousands of
    data points, and the decay in data spacing is more prominant. The
    relatively small reduction in $k$ along with the decrease in
    $\sigma_d$ explain the minimal reduction in error seen by Delaunay
    in Fig. \ref{fig:convergence-20d}.}
  \label{fig:data-spacing}
\end{figure}
}{}

\changed{The results achieved for both $d = 2$ and $d = 20$ align with
  the theoretical error bound as can be seen in Figure
  \ref{fig:data-spacing}. In low dimension, thousands of data points
  meaningfully reduce the sparsity of the approximation problem.
  However, in higher dimension it takes many more points to achieve a
  reduction in data sparsity while simultaneously being more difficult
  to produce well-\color{red}conditioned simplices. This evidences the inherent
  challenge of data sparsity in high dimension approximation
  problems. The analytic results presented are not caused by the
  chosen test function, but rather the exponential increase in
  complexity that accompanies increased dimension.}{}

In summary, the regime of \changed{signal-to-noise ratios that 
  result in competitively accurate interpolants}{} is \changed{}{much}
greater for moderate to high dimensional problems \changed{due to
  increased sparsity}{, even for functions with random noise
  incorporated into their evaluation}. Acknowledging the viability of
interpolation for problems of moderate dimension, the next section
will consider real-world problems of similar proportion (thousands of
examples in tens of dimensions).

@@

In addition to the figures displaying approximation results for each
data set, tables of accompanying numerical results \changed{including
  timings and exact prediction errors}{} are located in the Appendix
(Section \ref{sec:appendix}). All of the test data sets capture
underlying functions that are almost certainly stochastic. As
described in Section \ref{sec:introduction}, regression techniques
appear most appropriate for these problems. However, typically data
grows exponentially more sparse with increasing dimension. Given that
sparse data regressors tend towards interpolation and, as demonstrated
in Section \ref{sec:analytic}, interpolants produce similar (if not
identical) results, it is presumed that interpolants are equally
viable approximation techniques on these problems.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of forest fire area burned under recorded weather
    conditions. The data is presented on a $\ln$ scale because
    most values are small with exponentially fewer fires on record
    that burn large areas.}
  \label{fig:hist-forest-fire}

  \vspace{.3cm}

  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the amount of area
    that would be burned given environment conditions. $10$-fold cross
    validation as described in the beginning of Section \ref{sec:data}
    is used to evaluated each algorithm. This results in exactly one
    prediction from each algorithm for each data point. These boxes
    depict the median (middle bar), median $95\%$ confidence interval
    (cones), quartiles (box edges), fences at $3/2$ interquartile
    range (whiskers), and outliers (dots) of absolute prediction error
    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
    errors are presented on a $\ln$ scale. The numerical data
    corresponding to this figure is provided in Table
    \ref{table:error-forest-fire} in the Appendix.}
  \label{fig:error-forest-fire}
\end{figure}
